Observations from Parallelising Three Maximum Common (Connected) Subgraph Algorithms

### 2017-11-26 paper/paper.tex
Patrick's comments

There are three competitive approaches for the maximum common subgraph problem, each being the
strongest on certain classes of instance. The first involves a reduction to the maximum clique
problem, whilst the other two approaches have a constraint programming feel to them.

@@

McCreesh et al.\ \cite{o:McCreeshPT17} observe that due to the special structure of the maximum
common subgraph problem, the following property holds throughout the search process using the
constraint programming model: any two variables either have domains with no values in common (with
the possible exception of $\bot$), or have identical domains. The McSplit algorithm exploits this
property. It explores essentially the same search tree as the basic forward-checking constraint
programming model, but using different supporting algorithms and data structures.  Rather than
storing a domain for each vertex in the pattern graph, equivalence classes of vertices in both
graphs are stored in a special data structure which is modified in-place and restored upon
backtracking. This enables fast propagation of the constraints and smaller memory requirements. In
addition, this data structure enables stronger branching heuristics to be calculated cheaply. The
McSplit algorithm effectively dominates conventional constraint programming approaches, being
consistently over an order of magnitude faster.

@@

A different take on the constraint programming model is presented by
Hoffmann et al.\ \cite{DBLP:conf/aaai/HoffmannMR17}. They approach maximum common subgraph via the
subgraph isomorphism problem, asking the question ``if a pattern graph cannot be found in the
target, how much of the pattern graph can be found?''. The $k{\downarrow}$ algorithm tries to solve
the subgraph isomorphism problem first for $k=0$ (asking whether the whole pattern graph can be found in the
target). Should that not be satisfiable, it tries to solve the problem for $k=1$ (one vertex cannot
be matched), and should that also not be satisfiable, it iteratively increases $k$ until the result
is satisfiable. This approach exploits strong invariants using paths and the degrees of vertices to
prune large portions of the search space.

@@

We stress that these anomalies are due to changes in the amount of work done, and are not due to
work balance problems (although work balance is \emph{also} unusually difficult for this problem).
Anomalies can have a very strong effect on these algorithms, and we will therefore try to mitigate
them as far as possible. In the evaluation of their ``embarrassingly parallel search'' technique,
Malapert et al.\ \cite{DBLP:journals/jair/MalapertRR16} ``consider unsatisfiable, enumeration and
optimization [problem] instances'', and ``ignore the problem of finding a first feasible solution
because the parallel speedup can be completely uncorrelated to the number of workers, making the
results hard to analyze''. They do ``consider optimization problems for which the same variability
can be observed, but at a lesser extent because the optimality proof is required''. Unfortunately,
many of the instances we consider behave more like decision problem instances than optimisation
instances: due to the combination of a low solution density, good value-ordering heuristics, and a
strong bound function in cases where the optimal solution is relatively large, it is often the case
that the runtime is determined almost entirely by how long it takes to find an optimal solution,
with the proof of optimality being nearly trivial.  Indeed, attempts to parallelise the basic
constraint programming model by static decomposition have had limited success
\cite{DBLP:conf/ictai/MinotNS15}.

>>>>

There are three competitive approaches for the maximum common subgraph problem, each being the
strongest on certain classes of instance. The first involves a reduction to the maximum clique
problem, whilst the other two approaches are inspired by constraint programming.

@@

McCreesh et al.\ \cite{o:McCreeshPT17} observe that due to the special structure of the maximum
common subgraph problem, the following property holds throughout the search process using the
constraint programming model: any two variables either have domains with no values in common (with
the possible exception of $\bot$), or have identical domains. The McSplit algorithm exploits this
property. It explores essentially the same search tree as the basic forward-checking constraint
programming approach, but using different supporting algorithms and data structures.  Rather than
storing a domain for each vertex in the pattern graph, equivalence classes of vertices in both
graphs are stored in a special data structure which is modified in-place and restored upon
backtracking. This enables fast propagation of the constraints and smaller memory requirements. In
addition, this data structure enables stronger branching heuristics to be calculated cheaply. The
McSplit algorithm effectively dominates conventional constraint programming approaches, being
consistently over an order of magnitude faster.

@@

A different take on the constraint programming approach is presented by
Hoffmann et al.\ \cite{DBLP:conf/aaai/HoffmannMR17}. They approach maximum common subgraph via the
subgraph isomorphism problem, asking the question ``if a pattern graph cannot be found in the
target, how much of the pattern graph can be found?''. The $k{\downarrow}$ algorithm tries to solve
the subgraph isomorphism problem first for $k=0$ (asking whether the whole pattern graph can be found in the
target). Should that not be satisfiable, it tries to solve the problem for $k=1$ (one vertex cannot
be matched), and should that also not be satisfiable, it iteratively increases $k$ until the result
is satisfiable. This approach exploits strong invariants using paths and the degrees of vertices to
prune large portions of the search space.

@@

We stress that these anomalies are due to changes in the amount of work done, and are not due to
work balance problems (although work balance is \emph{also} unusually difficult for this problem).
Anomalies can have a very strong effect on these algorithms, and we will therefore try to mitigate
them as far as possible. In the evaluation of their ``embarrassingly parallel search'' technique,
Malapert et al.\ \cite{DBLP:journals/jair/MalapertRR16} ``consider unsatisfiable, enumeration and
optimization [problem] instances'', and ``ignore the problem of finding a first feasible solution
because the parallel speedup can be completely uncorrelated to the number of workers, making the
results hard to analyze''. They do ``consider optimization problems for which the same variability
can be observed, but at a lesser extent because the optimality proof is required''. Unfortunately,
many of the instances we consider behave more like decision problem instances than optimisation
instances: due to the combination of a low solution density, good value-ordering heuristics, and a
strong bound function in cases where the optimal solution is relatively large, it is often the case
that the runtime is determined almost entirely by how long it takes to find an optimal solution,
with the proof of optimality being nearly trivial.  Indeed, attempts to parallelise the basic
constraint programming approach by static decomposition have had limited success
\cite{DBLP:conf/ictai/MinotNS15}.
